Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2025-01-19 demystify.tex
updates

I will also let the student know if I cannot write a strong letter for them (e.g., I don't know them that well), and suggest they find someone else.  If they insist, then I will write for them. While I try to say something positive, e.g., the student is hardworking, or that I heard they are working on a research project, etc, the letter will still be short and weak (\autoref{sec:generic-letters}). Usually it takes me about \emph{5--10} minutes to write such a letter.

@@

\section{GREs Are Optional and Do Not Matter}\label{sec:gre}

@@

@@

\subsection{How Applications are Evaluate}

@@

>>>>

I will also let the student know if I cannot write a strong letter for them (e.g., I don't know them that well), and suggest they find someone else.  If they insist, then I will write for them. While I try to say something positive, e.g., the student is hardworking and receive good grades, the letter will still be short and weak (\autoref{sec:generic-letters}). Usually it takes me about \emph{5--10} minutes to write such a ``weak'' letter.

@@

\section{GREs Are Optional and Do Not Matter for PhD Admissions}\label{sec:gre}

@@

Note that while GRE is not important for CS PhD admission, it might be required or important for MS admission (\autoref{chap:ms}). This is because MS programs are more course-based and thus care more about grades and standardized tests.

@@

\subsection{How Applications are Evaluated}

@@

\section{DoD NDSEG}

The DoD National Defense Science and Engineering Graduate (NDSEG) is another prestigious fellowship supporting \emph{U.S. citizens} pursuing advanced degrees in STEM fields critical to national defense (such as CS).
It provides up to three years of funding, including a  stipend, tuition coverage, and health insurance.

\subsection{Evaluation}
Unlike the NSF GRFP, for NDSEG reviewers \emph{do not} see your LoRs, i.e., we do not know who wrote them or what they wrote. Moreover, we only see \emph{redacted versions} of your submitted materials (you are asked to provide a full and redacted version of your research plan, personal essays, and CV). Thus we do not know what schools your went to, where you did your internships, your publications or research experience. We can see your current transcript, but not your past ones.

Thus, reviewers mainly get to know you and evaluate your application by reading your research plan and personal essays. The reason is for fairness and avoiding bias based on the school you went to, places you published at, your LoR writers, etc.  It could be that external reviewers like myself help with the first cut based on the redacted materials, and then the final committee makes the final decision based on the full materials.

\begin{commentbox}[Vu:]
I am used to reading full applications and thus find it a bit difficult to evaluate NDSEG applications because I do not have the full picture.  For example, a student mentioned that they have published in the topic they proposed but redacted all information about the publication, so I have no idea where they published, what the paper is about, etc.  
It is also difficult to evaluate the academic background of the student, e.g., some student just started grad school and so their transcript is essentially blank, and I do not know where they went to undergrad, etc. It is also annoying to see CVs with so many blanked out information. They might as well should just submit a blank CV one. 
\\

In the beginning I find it surprising that I don't get to see the LoRs, but I understand the reasoning behind it and gradually think it is a good idea. As mentioned above, LoR writers, especially those from top schools, often overrate their students and thus can bias the evaluation. So I don't find I am missing much by not seeing the LoRs.
\end{commentbox}

\subsection{Common Pitfalls and Tips}
Based on my experience reviewing NDSEG applications, here are some common pitfalls to avoid:
\begin{itemize}

    \item Customize your research to DoD, e.g., through specific project from DARPA or ONR. 
    Make this prominent, e.g., in the introduction or in its own section.  This is different from NSF GRFP where you want broader impacts to society.

    \item Don't make up things or exaggerate your research.  Reviewers are experts in the field and can easily spot technical inaccuracies or ideas that have already been studied. Understandbly many applicants are new, but you need to do proper research on what you propose. In particular, you should aim to have (i) clear problem statement, (ii) why it is important (esp. to DoD), (iii) what has been done and their challenges/limitations, and (iv) what you plan to do to address the challenges.  Moreover, research is often quite technical and ambitious, so you want to show some preliminary work, e.g., a small experiment by hand on some small examples, to convince reviewers that you have thought about the problem and have some initial results.\\

    Unlike GRFP where we have more things to evaluate, in NDSEG we mainly evaluate you based on your personally essays and especially research plan. So you want to make sure your research plan is well thought out and convincing.

    \item Personal stories and essays should be unique and interesting.  Many personal stories are quite generic and do not tell much about the student. For example, being an URM, involving in a robotics club or mentoring underrepresented students are quite common and do not tell much.\\

    What you want is to pick a few thing and provide concrete examples and details.  For example, provide a concrete experience in being minority and how it has affected you, talk about what you have accomplished in your robotics club, e.g., you built a robot that won a competition, or you mentored a specific URM student who then went on to do great things.  Just a small, interesting detail can make your essay stand out.

    \item Not using all allowed pages. Many research plans use only 1/2 or 2/3 of the number of allowed pages, which themselves are already quite short. Such documents are often not detailed enough and do not provide enough information for reviewers to understand and evaluate the research plan.  So use all allowed pages to provide enough details about your research plan. This is especially important as we do not know much about you due to redacted materials.

\item 
personal essay / stories are written in text box and so do not llook very professional and doesn’t seem to be heavily weighted compared to research plan (e.g., very few questions asked about the personal stories compared to questions asked on research plan).

- use latex,  do not write in word and especially use left justification  (see how CS and engineer people publish papers)  
- If you have an advisor or Lor writer, ask them to look at your writing/essays .  This is even suggested directly by NSF and DoD.  You are new to the field and what your ideas might not be accurat or have already been explored — let them help you here.  While you can BS to PhD adcom because there you don’t really write a research plan,  but for these fellowships you do and reviewers are all experts in the field and would look at these carefully.
